The \dealii Library, Version 9.1

### 2019-06-05 9.1/paper.tex
add a reference to the ginkgo project
Same problem here: I haven't found a citable publication
Changed my affiliation to Swansea University where most of the work was undertaken and added support information to the acknowledgements section.

\author[9]{Ross~Maguire~Kynch}
\affil[9]{Renishaw Plc,
  Miskin, CF72 8XY, Wales, UK.
  {\texttt{rkynch@gmail.com}}}

@@

\dealii version 9.1.0 was released May 21, 2019.
This paper provides an
overview of the new features of this major release and serves as a citable
reference for the \dealii software library version 9.1. \dealii is an
object-oriented finite element library used around the world in the
development of finite element solvers. It is available for free under the
GNU Lesser General Public License (LGPL). Downloads are available at
\url{https://www.dealii.org/} and \url{https://github.com/dealii/dealii}.

The major changes of this release are:
%
\begin{itemize}
  \item Improved support for automatic differentiation (see
    Section~\ref{subsec:ad}),
  \item Dedicated support for symbolic algebra (see
    Section~\ref{subsec:sd}),
  \item Full support for $hp$~adaptivity in parallel computations (see
    Section~\ref{subsec:hp}),
  \item Interfaces to the HDF5 file format and libraries (see
    Section~\ref{subsec:hdf5}),
  \item Significantly extended GPU support (see Section~\ref{subsec:gpu}),
  \item Parallel geometric multigrid (GMG) improvements (see
    \cite{ClevengerHeisterKanschatKronbichler2019} and
    Section~\ref{subsec:gmg}),
  \item Four new tutorial programs (step-61, step-62, step-63, step-64),
    as well as one new code gallery program (see
    Section~\ref{subsec:steps}).
\end{itemize}
%
The major changes are discussed in detail in Section~\ref{sec:major}. There
are a number of other noteworthy changes in the current \dealii{} release
that we briefly outline in the remainder of this section:
%
\begin{itemize}
\item
  The release contains bug fixes and performance improvements of the
  matrix-free framework and related geometric multigrid solvers. In
  particular, the implementation of the Chebyshev iteration, an often used
  smoother in the matrix-free context, has been revised to reduce vector
  accesses. This speeds up matrix-free multigrid solvers by up to 10-15\%
  on geometries with affine (parallelogram and parallelpiped) cells, and up
  to 5\% on geometries with cells bounded by curved edges and faces.

\item
  Various variants of geometric multigrid solvers and matrix-free
  implementations were run on up to 304,128 MPI ranks during the acceptance
  phase of the SuperMUC-NG supercomputer, verifying the scalability of our
  implementations. Some performance bottlenecks on distributed
  triangulation with more than 100k ranks have been fixed.

@@

In the previous release, numerous classes that are used to assemble linear systems
and right hand sides, as well those used to define constitutive laws, were given
full support for ``white-listed'' automatically differentiable (AD) number types
from the ADOL-C and Sacado libraries. This included the classes that
represent the local contributions of one cell to the global linear
system (i.e., \texttt{FullMatrix} and \texttt{Vector}) as well as the
classes commonly used for the computations at individual quadrature
points (primarily the \texttt{Tensor} and \texttt{SymmetricTensor}
classes of various ranks).
In the current release we have provided a unified interface to these
AD libraries, focussing on two specific use contexts:
\begin{enumerate}
\item The construction and linearization of finite element residuals; and
\item The construction and linearization of constitutive model kinetic variables.
\end{enumerate}

In the first context, the finite element degrees of freedom are considered the
independent variables. From these primitives, the \texttt{EnergyFunctional} helper
class in the namespace \texttt{Differentiation::AD} may be used to compute both the
residual and its linearization by directly defining the contribution to the
(twice differentiated) scalar total energy functional from each cell. Similarly,
the \texttt{ResidualLinearization} class requires the (once-differentiated) finite
element residual to be defined on a per cell basis,  and this contribution is
automatically linearized.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Interfaces to the HDF5 file format and libraries}
\label{subsec:hdf5}

\marginpar{Missing}

@@

GPU support was significantly extended for the current release:
\begin{itemize}
  \item The \texttt{CUDAWrappers::PreconditionILU} and \texttt{CUDAWrappers::PreconditionIC}
    classes can now be used for preconditioning \texttt{CUDAWrappers::SparseMatrix} objects.
  \item \texttt{LinearAlgebra::distributed::Vector}: the MPI-parallel vector
    class has gained a second template argument \texttt{MemorySpace} which can
    either be \texttt{Host} or \texttt{CUDA}. In the latter case, the data
    resides in GPU memory. By default, the template parameter is
    \texttt{Host} and the behavior is unchanged compared to previous versions.
    When using CUDA, the ghost exchange can be performed either by first copying
    the relevant data to the host, performing MPI communication, and finally
    move the data to the device or, if CUDA-aware MPI is available, by
    performing MPI communication directly between GPUs.
  \item Constrained degrees of freedom: the matrix-free framework now
    supports constrained degrees of freedom. The implementation is based on
    the algorithms described in \cite{ljungkvist2017}. With this addition,
    both Dirichlet boundary conditions and the constraints arising from
    adaptively refined meshes can be imposed within the matrix-free
    framework.
    \textcolor{red}{%
    \sout{The only restriction is that for two-dimensional meshes the
    finite element degree must be odd. There is no such restriction in
    three dimensions.}
    \emph{%
    WB: Can we explain this restriction? It seems odd (pun intended)...
    BT: No, we cannot. I have spent at least two full weeks trying to
    understand why it doesn't work, i.e., gives the wrong result, but I
    don't know. We decided at the time that even with the current
    limitation this was a useful addition.
    MM: I suggest that we then simply not mention the limitation and treat
    it as a known bug. As it is currently worded it can be easily
    interpreted as a fundamental limitation, i.\,e. ``FE with even degree
    are not admissible.''}}

@@

\textcolor{red}{Missing: performance optimizations, new features, largest
runs now at XYZ (Kronbichler?)}

@@

>>>>

\author[9]{Ross~Maguire~Kynch}
\affil[9]{Zienkiewicz Centre for Computational Engineering,
  College of Engineering, Swansea University,
  Bay Campus, Fabian Way, Swansea SA1 8EN, Wales, UK.
  {\texttt{rkynch@gmail.com}}}

@@

\dealii version 9.1.0 was released May 21, 2019.
This paper provides an
overview of the new features of this release and serves as a citable
reference for the \dealii software library version 9.1. \dealii is an
object-oriented finite element library used around the world in the
development of finite element solvers. It is available for free under the
GNU Lesser General Public License (LGPL). Downloads are available at
\url{https://www.dealii.org/} and \url{https://github.com/dealii/dealii}.

The major changes of this release are:
%
\begin{itemize}
  \item Improved support for automatic differentiation (see
    Section~\ref{subsec:ad}),
  \item Dedicated support for symbolic algebra (see
    Section~\ref{subsec:sd}),
  \item Full support for $hp$~adaptivity in parallel computations (see
    Section~\ref{subsec:hp}),
  \item An interface to the HDF5 file format and libraries (see
    Section~\ref{subsec:hdf5}),
  \item Significantly extended GPU support (see Section~\ref{subsec:gpu}),
  \item Parallel geometric multigrid (GMG) improvements (see
    \cite{ClevengerHeisterKanschatKronbichler2019} and
    Section~\ref{subsec:gmg}),
  \item Four new tutorial programs (step-61, step-62, step-63, step-64),
    as well as one new code gallery program (see
    Section~\ref{subsec:steps}).
\end{itemize}
%
The major changes are discussed in detail in Section~\ref{sec:major}. There
are a number of other noteworthy changes in the current \dealii{} release
that we briefly outline in the remainder of this section:
%
\begin{itemize}
\item
  The release contains a number of performance improvements and bug fixes for
  the matrix-free framework. One notable improvement is the support for
  renumbering of degrees of freedom within the cells for discontinuous
  elements, avoiding some reshuffling operations across the SIMD lanes
  with vectorization over several cells and faces, which is especially
  useful on processors with AVX-512 vectorization (8 doubles), speeding up
  operations by up to 10\%. Secondly, the strategy for the most efficient
  tensor product evaluators according to the performance analysis of
  \cite{KronbichlerKormann2019} in the context of more quadrature points than
  shape functions has been revised for better performance.

@@

In the previous release, numerous classes that are used to assemble linear systems
and right hand sides, as well those used to define constitutive laws, were given
full support for ``white-listed'' automatically differentiable (AD) number types
from the ADOL-C and Sacado libraries. This included the classes that
represent the local contributions of one cell to the global linear
system (i.e., \texttt{FullMatrix} and \texttt{Vector}) as well as the
classes commonly used for the computations at individual quadrature
points (primarily the \texttt{Tensor} and \texttt{SymmetricTensor}
classes of various ranks).
In the current release we have provided a unified interface to these
AD libraries, focusing on two specific use contexts:
\begin{enumerate}
\item The construction and linearization of finite element residuals; and
\item The construction and linearization of constitutive model kinetic variables.
\end{enumerate}

In the first context, the finite element degrees of freedom are considered the
independent variables. From these primitives, the \texttt{EnergyFunctional} helper
class in the namespace \texttt{Differentiation::AD} may be used to compute both the
residual and its linearization by directly defining the contribution to the
(twice differentiated) scalar total energy functional from each cell. Similarly,
the \texttt{ResidualLinearization} class requires the (once-differentiated) finite
element residual to be defined on a per-cell basis,  and this contribution is
automatically linearized.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Interface to the HDF5 file format and libraries}
\label{subsec:hdf5}
\texttt{HDF5} is an open-source library and file format designed to store large amounts of data.
The \texttt{HDF5} format is specially tailored for high volume parallel I/O operations using MPI.
\texttt{HDF5} files are self-describing and allow complex data relationships and a large variety
of datatypes. In addition, the \texttt{HDF5} format is designed for long-term preservation of
data; a \texttt{HDF5} file created by an HPC system can be easily read by a commodity laptop.

\dealii{}'s new  \texttt{HDF5} interface allows to write \texttt{HDF5} files and manipulate
datasets, groups, and attributes in serial and in parallel using MPI. The
\texttt{HDF5} MPI library calls that modify the structure of the file are
always collective, whereas writing and reading raw data in a dataset can be done
independently or collectively. In the \dealii{}'s \texttt{HDF5} interface all the calls are
set to collective in order to maximize performance. This means that all the MPI
processes have to contribute to every single call, even if they don't have data to write.
We have added the following classes to the \dealii{}'s \texttt{HDF5} namespace.
\begin{itemize}
\item The new \texttt{HDF5::File} class can be used to open and create \texttt{HDF5} files
in serial or in parallel.
\item The new \texttt{HDF5::Group} class can be used to open and create groups.
 \texttt{HDF5} files have a tree structure. The root contains groups and
the groups can contain other groups.
\item The new \texttt{HDF5::DataSet} class can be used to open and create datasets which
can be placed inside groups or at the root of the \texttt{HDF5} file. A dataset can be a
vector, a matrix, or a tensor. It is possible to read and write in the dataset using
hyperslabs or unordered data. Hyperslabs are portions of datasets
which can be a contiguous collection of points in a dataset, or a regular
pattern of points or blocks in a datataset.
\item It is possible to define attributes in the \texttt{HDF5::File}, \texttt{HDF5::Group} and
\texttt{HDF5::DataSet} classes. An attribute is a small metadata such as a number or a string.
Attributes are commonly used to store simulation parameters.
\end{itemize}
As we have shown in \texttt{step-62}, the \dealii{}'s \texttt{HDF5} interface can be easily used
to exchange data with python and jupyter notebooks.

@@

GPU support was significantly extended for the current release:
\begin{itemize}
  \item The \texttt{CUDAWrappers::PreconditionILU} and \texttt{CUDAWrappers::PreconditionIC}
    classes can now be used for preconditioning \texttt{CUDAWrappers::SparseMatrix} objects.
  \item \texttt{LinearAlgebra::distributed::Vector}: the MPI-parallel vector
    class has gained a second template argument \texttt{MemorySpace} which can
    either be \texttt{Host} or \texttt{CUDA}. In the latter case, the data
    resides in GPU memory. By default, the template parameter is
    \texttt{Host} and the behavior is unchanged compared to previous versions.
    When using CUDA, the ghost exchange can be performed either by first copying
    the relevant data to the host, performing MPI communication, and finally
    move the data to the device or, if CUDA-aware MPI is available, by
    performing MPI communication directly between GPUs.
  \item Constrained degrees of freedom: the matrix-free framework now
    supports constrained degrees of freedom. The implementation is based on
    the algorithms described in \cite{ljungkvist2017}. With this addition,
    both Dirichlet boundary conditions and the constraints arising from
    adaptively refined meshes can be imposed within the matrix-free
    framework.
%     \textcolor{red}{%
%     \sout{The only restriction is that for two-dimensional meshes the
%     finite element degree must be odd. There is no such restriction in
%     three dimensions.}
%     \emph{%
%     WB: Can we explain this restriction? It seems odd (pun intended)...
%     BT: No, we cannot. I have spent at least two full weeks trying to
%     understand why it doesn't work, i.e., gives the wrong result, but I
%     don't know. We decided at the time that even with the current
%     limitation this was a useful addition.
%     MM: I suggest that we then simply not mention the limitation and treat
%     it as a known bug. As it is currently worded it can be easily
%     interpreted as a fundamental limitation, i.\,e. ``FE with even degree
%     are not admissible.''}}

@@

For the 9.1 release, the geometric multigrid facilities have been extended and
revised for performance. The geometric multigrid algorithms for uniform and
adaptively refined meshes in \dealii are based on so-called local coarsening,
i.e., smoothening is applied level-per-level, skipping parts of the domain
where the mesh is not as refined. The algorithm for the assignment of the
owner on level cells and the implications on load balancing have been analyzed
in detail in \cite{ClevengerHeisterKanschatKronbichler2019}. While most of the
functionality was already available since the 8.5 release of \dealii presented
in \cite{dealII85}, several components have been finalized, such as the
support for certain renumbering algorithms that are beneficial for matrix-free
execution, and interfaces that allow the combination with matrix-free GPU
computations according to \cite{KronbichlerLjungkvist2019}.

A number of data structures and implementations in \dealii have been adapted
to ensure scalability of the matrix-free algorithms and geometric multigrid
infrastructure on more than 100,000 MPI ranks. Geometric multigrid solver for
the Poisson equation as described in \cite{KronbichlerWall2018} have been used
as performance tests during the acceptance phase of the SuperMUC-NG
supercomputer in Garching, Germany. Scaling tests have been performed on up to
the full machine with 304,128 cores of the Intel Xeon Skylake architecture and
an arithmetic performance of around 5 PFlop/s for a geometric multigrid solver
with polynomials of degree 4 has been reached. Compared to the official
LINPACK performance of the machine of 19.5 PFlop/s (the machine is listed on
position 8 of the top-500 list of November 2018), this can be considered an
extremely good value for a PDE solver which have classically only reached a
few percent of the LINPACK performance, and given the optimization steps
contained in the matrix-free module of \dealii that aim to minimize the time
to solution rather than maximize the number of floating point operations. The
largest Poisson problem that has been solved on 304k cores contained 2.15
trillion unknowns (or 7.1 million unknowns per MPI rank) and was solved in 3.5
seconds. Also, CFD production runs with up to $10^{11}$ unknowns and $10^5$
time steps have been completed in less than seven hours, demonstrating the
capabilities of \dealii for large-scale parallel computations. The scaling tests
also revealed several relatively expensive operations in the setup of the
multigrid unknowns in \dealii's \texttt{DoFHandler} and \texttt{MGTransfer}
classes. While a few bottlenecks have already been resolved, we plan several
further improvements in setup times for the next release.

Furthermore, the implementation of the Chebyshev iteration, \dealii's most
popular smoother in the matrix-free context, has been revised to reduce the
vector accesses. The vector operations have become an increasing bottleneck
due to the level of optimization available for the operator evaluation with
our matrix-free framework \cite{KronbichlerKormann2019}, especially on newer
CPUs which can perform a large number of arithmetic operations per byte loaded
from main memory. This speeds up matrix-free multigrid solvers by up to
10--15\% on geometries with affine (parallelogram and parallelpiped) cells,
and up to 5\% on deformed cells.

@@

M.~Kronbichler was also supported by the Bayerisches Kompetenznetzwerk
f\"ur Technisch-Wissen\-schaft\-li\-ches Hoch- und H\"ochstleistungsrechnen
(KONWIHR) in the context of the project
``Performance tuning of high-order discontinuous Galerkin solvers for
SuperMUC-NG''.

R.~M.~Kynch was supported by the Engineering and Physical Science Research
Council (EPSRC) UK through grant EP/K023950/1 while working at Swansea
University, UK 2013-2015 where the majority of his contribution was
undertaken.
